\bf Supplement to ``On Devon Allen's Disqualification at the 2022 World Track and Field Championships''

### 2024-11-30 Manuscript/manuscript.tex
touch up; removed addressed comments

The rest of this paper is organized as follows. Section~\ref{sec:Data}
describes the data collection and datasets used for the two
objectives. Section~\ref{sec:rank} outlines the rank-based comparison
methods for objective~1, rank-based comparison for clustered data with
subunit-level grouping to analyze differences in RTs across different
competitions. Section~\ref{sec:gamlss} details the methods for
objective~2, the GG model with random effects to evaluate the appropriatenes
of the 0.1-second threshold. Section~\ref{sec:Results} presents the
results of both objectives and discusses the implications.
Finally, Section~\ref{sec:concludingremarks} highlights the
paper’s impact, limitations, and potential for future research.
All data and code for our analysis are provided in the supplementary
materials.

@@

This study focuses on RTs from sprint events where reaction
time plays a critical role. Specifically, we included data from men’s
100-meter dashes and 110-meter hurdles, as well as women’s 100-meter
dashes and 100-meter hurdles. Statistical tests showed no significant
differences in RTs between these events, supporting their
inclusion in a unified analysis. However, we excluded data from
200-meter dashes and longer events, as their RTs were found
to be significantly different. This difference is not unexpected, as
RT is generally less critical in longer sprint events
compared to shorter distances where explosive starts are paramount.
Negative RTs were excluded from the analysis, while positive
disqualified RTs were included due to their low frequency. A
sensitivity analysis regarding the inclusion of positive disqualified
RTs is provided in the Supplementary Material.

@@

The data for evaluating the appropriateness of the 0.1-second
threshold was obtained from World Athletics and covers the men's
110-meter hurdles and 100-meter dashes from 1999 to 2023. Due to
possible gender differences \citep{babicc2009reaction,
  lipps2011implications}, data for women's 100-meter hurdles and
100-meter dashes and their analyses were relegated to the
supplementary material. We focus on the RTs recorded during
semifinal and final heats only, as RTs from preliminary
heats are often not as fast as those in later heats
\citep[e.g.,][]{collet1999strategic, tonnessen2013reaction,
  brosnan2017effects, zhang2021correlation}. For analysis purposes, we
pooled RTs from semifinal and final heats to increase sample
size, which is particularly important for years with limited final heat
observations. For example, in 2022, only five data points were available
from the final heat due to two disqualifications and one athlete not
competing. Unless otherwise noted, this pooled dataset forms the basis
for our analyses for Objective 2. Additionally, we consider datasets
that exclude 2022 to assess how our findings might differ when excluding
this year of interest.

@@

Table~\ref{tab:Clusrankresults} presents the results from both permutation
and asymptotic rank-based tests for these six comparisons. The tests reveal
that the national versus international comparisons yielded the lowest
p-values, indicating stronger evidence of faster RTs at the
2022 World Championships compared to the 2022 national competitions. 
\eds{This does not appear true based on the data currently in Table 1. 
2022-2023 seem to have the lowest p-values.  Which is correct?}
For both men and women, permutation test results were highly significant,
providing substantial evidence that athletes competing at the 2022 World
Championships had significantly faster RTs than they did just
a few months earlier at national-level events. The comparisons between the
2019 and 2022 World Championships and between the 2022 and 2023 World
Championships also produced significant results. Collectively,
these results suggest that athletes consistently achieved faster reaction
times at the 2022 World Championships compared to other competitions. This
finding supports the hypothesis that the 2022 World Championships presented
conditions conducive to faster RTs, potentially due to systematic
or environmental factors.

@@

The results reported in this subsection are from men's data only
because our investigation found significant gender difference.
Results for women's data are reported in Section~2 of the
Supplementary Material. The fitted parameters of the GG distribution in the
GAMLSS framework in Equations~\eqref{eq:gg}--\eqref{eq:sigma} are
summarized in Table~\ref{tab:ggfit}. Results obtained from both
excluding and including 2022 data are reported. The fixed-effect
parameters include $\beta_0$, $\gamma_0$, and $\nu$, corresponding to
the intercept of the log-location, log-scale, and shape of the GG
distribution, respectively. Random effects account for variability at
the venue level $\tau_v$ on the log-scale of the $\mu$ parameter
and at the heat level $\tau_h$ on the log-scale of the $\sigma$
parameter in the density in Equation~\eqref{eq:gg}. The variance
of the venue-level random effect is smaller than the heat-level random
effect variance, suggesting that heat-level variability in the scale
parameter is substantial, though on the dispersion parameter. When the
2022 data is included, all parameter estimates remain stable except
the standard deviation of the venue-level random-effect variance,
which increases from 0.043 to 0.058 . 
\eds{are you reporting estimated variances or SDs of the random effects here and 
in the Table?  In the table, we have $\tau_v$ and $\tau_h$ (SD) labels, not 
$\tau_v^2$ and $\tau_h^2$ (variances).  Please check and correct as needed.}  
These results
highlight that RTs are influenced by both venue and heat-level
factors, and that the inclusion of 2022 introduces greater venue-level
variability, likely due to systematic differences in RTs that year.

@@

In addressing the second question, our analyses of the GAMLSS model
reveal that the 0.1-second RT barrier, while rare, may not
be as extraordinary  as traditionally assumed. For men,
Table~\ref{tab:Sim_time} shows that reaction  times below 0.1 seconds
occur with a probability of approximately one in 362 starts  when
including the 2022 data. Lowering the threshold to 0.08 seconds
drastically  reduces this likelihood, supporting the idea that the
current barrier could be adjusted to reflect more realistic
probabilities of false starts. A similar pattern is observed for
women, as detailed in the Supplementary Material, where reaction
times below 0.1 seconds are exceedingly rare for the 100-meter dash
and 100-meter  hurdles. However, the uniformity of the 0.1-second
barrier for both men and women  is questionable, given numerous
studies documenting gender differences in reaction  times
\citep[e.g.,][]{lipps2011implications, babicc2009reaction,
  panoutsakopoulos2020gender}. These studies suggest that the current
threshold may  unfairly penalize men, for whom sub-0.1-second reaction
times are more probable. \citet{brosnan2017effects} advocate for
gender-specific barriers, a position that  aligns with our findings
and highlights the importance of tailoring thresholds to  biological
distinctions. \eds{is this still true in the updated analyses?}

@@

\section*{Supplementary Material}
\eds{when the supplement is complete, include an additional sentence here
that summarizes the additional analyses presented in the supplemental material}
The data and R code used for the analysis are available in a compressed file for
ease of reproducibility.

>>>>

The rest of this paper is organized as follows. Section~\ref{sec:Data}
describes the data collection and datasets used for the two
objectives. Section~\ref{sec:rank} outlines the rank-based comparison
methods for objective~1, rank-based comparison for clustered data with
subunit-level grouping to analyze differences in RTs across different
competitions. Section~\ref{sec:gamlss} details the methods for
objective~2, the GG model with random effects to evaluate the appropriatenes
of the 0.1-second threshold. Section~\ref{sec:Results} presents the
results of both objectives and discusses the implications.
Finally, Section~\ref{sec:concludingremarks} highlights the
paper’s impact, limitations, and potential for future research.
All data and code for our analysis are provided in the Supplementary
Materials.

@@

This study focuses on RTs from sprint events where reaction
time plays a critical role. Specifically, we included data from men’s
100-meter dashes and 110-meter hurdles, as well as women’s 100-meter
dashes and 100-meter hurdles. Statistical tests showed no significant
differences in RTs between these events, supporting their
inclusion in a unified analysis. However, we excluded data from
200-meter dashes and longer events, as their RTs were found
to be significantly different. This difference is not unexpected, as
RT is generally less critical in longer sprint events
compared to shorter distances where explosive starts are paramount.
Negative RTs were excluded from the analysis, while positive
disqualified RTs were included due to their low frequency and not
being obvious outliters. % A
% sensitivity analysis regarding the inclusion of positive disqualified
% RTs is provided in the Supplementary Material.

@@

The data for evaluating the appropriateness of the 0.1-second
threshold was obtained from World Athletics and covers the men's
110-meter hurdles and 100-meter dashes from 1999 to 2023. Due to
possible gender differences \citep{babicc2009reaction,
  lipps2011implications}, data for women's 100-meter hurdles and
100-meter dashes and their analyses were relegated to the
Supplementary Material. We focus on the RTs recorded during
semifinal and final heats only, as RTs from preliminary
heats are often not as fast as those in later heats
\citep[e.g.,][]{collet1999strategic, tonnessen2013reaction,
  brosnan2017effects, zhang2021correlation}. For analysis purposes, we
pooled RTs from semifinal and final heats to increase sample
size, which is particularly important for years with limited final heat
observations. For example, in 2022, only five data points were available
from the final heat due to two disqualifications and one athlete not
competing. Unless otherwise noted, this pooled dataset forms the basis
for our analyses for Objective 2. Additionally, we consider datasets
that exclude 2022 to assess how our findings might differ when excluding
this year of interest.

@@

Table~\ref{tab:Clusrankresults} presents the results from both permutation
and asymptotic rank-based tests for the six comparisons. All tests
yielded very small p-values, even after applying a Bonferroni
adjustment for multiple comparisons, indicating consistent evidence of
faster RTs at the 2022 World Championships relative to other
competitions. For both men and women, the national versus
international comparisons showed that RTs at the 2022 World
Championships were significantly faster than at national-level
competitions held earlier that year. Similarly, comparisons between
the 2019 and 2022 World Championships and between the 2022 and 2023
World Championships produced significant results, reinforcing the
observation that RTs at the 2022 World Championships were notably
faster. These findings support the hypothesis that conditions at the
2022 World Championships, whether systematic or
environmental, contributed to consistently faster RTs.

@@

The results reported in this subsection are from men's data only
because our investigation found significant gender difference.
Results for women's data are reported in Section~2 of the
Supplementary Material. The fitted parameters of the GG distribution in the
GAMLSS framework in Equations~\eqref{eq:gg}--\eqref{eq:sigma} are
summarized in Table~\ref{tab:ggfit}. Results obtained from both
excluding and including 2022 data are reported. The fixed-effect
parameters include $\beta_0$, $\gamma_0$, and $\nu$, corresponding to
the intercept of the log-location, log-scale, and shape of the GG
distribution, respectively. Random effects account for variability at
the venue level on the log-scale of the $\mu$ parameter
and at the heat level on the log-scale of the $\sigma$
parameter in the density in Equation~\eqref{eq:gg}. The variance
of the venue-level random effect is smaller than the heat-level random
effect variance, suggesting that heat-level variability in the scale
parameter is substantial, though on the dispersion parameter. When the
2022 data is included, all parameter estimates remain stable except
the standard deviation of the venue-level random-effect variance,
which increases from 0.043 to 0.058. These results
highlight that RTs are influenced by both venue and heat-level
factors, and that the inclusion of 2022 introduces greater venue-level
variability, likely due to systematic differences in RTs that year.

@@

In addressing the second question, our analyses of the GAMLSS model
reveal that the 0.1-second RT barrier, while rare, may not
be as extraordinary  as traditionally assumed. For men,
Table~\ref{tab:Sim_time} shows that reaction  times below 0.1 seconds
occur with a probability of approximately one in 362 starts  when
including the 2022 data. Lowering the threshold to 0.08 seconds
drastically  reduces this likelihood, supporting the idea that the
current barrier could be adjusted to reflect more realistic
probabilities of false starts. A similar pattern is observed for
women, as detailed in the Supplementary Material, where reaction
times below 0.1 seconds are exceedingly rare for the 100-meter dash
and 100-meter  hurdles. However, the uniformity of the 0.1-second
barrier for both men and women  is questionable, given numerous
studies documenting gender differences in reaction  times
\citep[e.g.,][]{lipps2011implications, babicc2009reaction,
  panoutsakopoulos2020gender}. These studies suggest that the current
threshold may  unfairly penalize men, for whom sub-0.1-second reaction
times are more probable. \citet{brosnan2017effects} advocate for
gender-specific barriers, a position that  aligns with our findings
and highlights the importance of tailoring thresholds to  biological
distinctions.

@@

\section*{Supplementary Material}
Additional results are summarized in a supplement for rank-based
comparison with pooled (men and women) data and for the GAMLSS results
for women data.
The data and R code used for the analysis are available in a compressed file for
ease of reproducibility.

### 2024-11-30 Manuscript/supp.tex
touch up; removed addressed comments

\jy{Keep the order consistent with the text; do the same in the code.}

As an alternative to the methods described in Section 3.1 of the main paper, it
is possible to combine all men's and women's reaction time (RT) data for each of 
the three competition
comparisons.  Thus we shrink our analyses from six to three, but each analysis
is roughly twice as big as previously.  For the 2019 versus 2022 comparison,
there were 258 RTs from 65 athletes and the asymptotic
test result was a p-value of $3.56 \cdot 10^{-8}$. For the 2022 national versus
international comparison, there were 160 RTs from 35
athletes and the asymptotic test result was a p-value of $1.94 \cdot 10^{-7}$.
For the 2023 versus 2022 comparison, there were 343 RTs
from 92 athletes and the asymptotic test result was a p-value of
$4.99 \cdot 10^{-12}$.  As a result of the larger sample sizes, the code was
very computationally expensive to run and thus we ran it on a smaller size.  All
three permutation tests had a p-value of $1 \cdot 10^{-5}$, which is the smallest
possible value given the permutation test size of $100,000$.  When taken together
with the asymptotic results, the results becoems very clear. These are all
highly significant test results that show substantial differences in average RT
for athletes competing at multiple championship-level events.

@@

We can repeat our reaction time barrier analysis described in Section 3.2, but
fit the same model to women's RT data from 2001 to 2023.  Once again we utilize
the power of the venue effect within $\mu$ and a heat effect within $\sigma$.
\eds{what is meant by `power' here?} 
The RT data for women is visualized in Figure~\ref{fig:WomensBoxplot}.
Similar to the men's data, RTs from 2022 appear lower than in other
years.

@@

We examined several models in the generalized Gamma model and found that the
best model for women involved a venue effect in $\mu$ and heat effect in
$\sigma$.  Thus the best model for men was also the best model for women.

@@

One of the most interesting aspects of Table~\ref{tab:womensfit} is that while
the venue effect standard deviation is nearly identical, the women's heat effect
standard deviation is smaller and $\nu$ is much larger.  If men's races have
more variability in times, it seems possible that in men's races that faster
athletes can influence other athletes to react faster in certain instances.
It is also interesting that the standard deviation of the venue effect is
nearly identical for both men's and women's data, indicating that the venue
effect is not only statistically significant but is consistent in strength
across men and women.

We can repeat the simulation methods described in the main paper to evaluate
the probability of an extreme reaction time.  Table~\ref{tab:Sim_prob_women}
compares the men's and women's results of observing reaction times less than
0.08, 0.09, and 0.1 seconds.  We find that across all three comparisons that
women have a lower probability of having a fast time. The results are slightly
different from those from the men's data, which echoes existing studies
reporting gender differences in RTs \citep[e.g.,][]{lipps2011implications,
babicc2009reaction, panoutsakopoulos2020gender}.

@@

Now we need to perform alter the interpretation of the analyis to determine
what a fair reaction time barrier for women should be.
Table~\ref{tab:Sim_time_women} shows the difference in men's and women's
suggested reaction time barrier.  What we see is that by World Athletics setting
the same standard for men and women, their rules are not fair, in that men are
more likely to be disqualified because they are more likely to react quickly.
One of the biggest takeaways of the reaction time barrier analysis is that the
current standards imposed by World Athletics are not grounded in statistical
analysis and that they should be.  It seems reasonable to either raise the
women's barrier so that they are disqualified at a similar rate as men or to
lower the men's barrier so they are disqualified at a similar rate as women.  But
having equal but not neccessarily fair standards does not seem reasonable when
this study's results echoed results of numerous other papers asking World
Athletics to change the standards.

@@

\section{Results from Data Excluding Positive Disqualified RTs}

An earlier iteration of the paper fit a model that did not include RTs from
athletes who were disqualified or did not finish but still registered a RT.  
However, it was ultimately decided to include these times to better estimate the 
left tail of the distribution and more accurately predict the probability of a 
low RT, as described in the main paper.  
We did not include negative RTs, however, as these represent a mistake 
of the runner for starting before
the gun is fired and are thus meaningless in our objective to
determine a fair RT barrier.  Not all of those disqualified were
disqualified because of breaking the 0.1 reaction time barrier; there are many
reasons why an athlete may be disqualified, with the most notable being failed
drug tests and lane violations.  Nonetheless, in this section,
we exclude all disqualified RTs to see their effect on the probability of an
extreme RT.  We fit an identical model to the generalized Gamma model
presented in Section 3.2 to examine differences results.

\begin{table}
  \centering
  \caption{Probabilities of observing RTs less than threshold 0.08,
  0.09, and 0.10 seconds based on the
    fitted GG GAMLSS model with both venue- and heat-level
random effects.}
  \begin{tabular}{c c c c}
   \toprule
   Data Set & Threshold 0.08 & Threshold 0.09 & Threshold 0.10  \\
   \midrule
   Without DQs & $4.93\cdot10^{-5}$ & $3.53\cdot10^{-4}$ &  $1.97\cdot10^{-3}$  \\
   With DQs & $6.84\cdot10^{-5}$ & $4.95\cdot10^{-4}$ & $2.76\cdot10^{-3}$ \\
   \bottomrule
  \end{tabular}
  \label{tab:DQSim_probability}
\end{table}

\jy{Men's data or pooled data?}
\eds{with or without 2022?}
\of{I dont know if we have time for this or if this fits the scope of what we
are working on.}

Table~\ref{tab:DQSim_probability} shows the effect of removing disqualificated
times from the analysis.  The probability of observing extreme RTs is lower when
we remove the 17 observations.

>>>>

As an alternative to the methods described in Section 3.1 of the main paper, it
is possible to combine all men's and women's reaction time (RT) data for each of 
the three competition
comparisons.  Thus we shrink our analyses from six to three, but each analysis
is roughly twice as big as previously.
For the 2022 national versus
international comparison, there were 160 RTs from 35
athletes and the asymptotic test result was a p-value of $1.94 \cdot 10^{-7}$.
For the 2019 versus 2022 comparison,
there were 258 RTs from 65 athletes and the asymptotic
test result was a p-value of $3.56 \cdot 10^{-8}$. 
For the 2023 versus 2022 comparison, there were 343 RTs
from 92 athletes and the asymptotic test result was a p-value of
$4.99 \cdot 10^{-12}$.  As a result of the larger sample sizes, the code was
very computationally expensive to run and thus we ran it on a smaller size.  All
three permutation tests had a p-value of $1 \cdot 10^{-5}$, which is the smallest
possible value given the permutation test size of $100,000$.  When taken together
with the asymptotic results, the results becoems very clear. These are all
highly significant test results that show substantial differences in average RT
for athletes competing at multiple championship-level events.

@@

We can repeat our reaction time barrier analysis described in Section 3.2, but
fit the same model to women's RT data from 2001 to 2023.
The RT data for women is visualized in Figure~\ref{fig:WomensBoxplot}.
Similar to the men's data, RTs from 2022 appear lower than in other
years. After removing one obvious outliter, which was a disqualfied
reaction time, the same model for men's data fits the women's data
reasonably well.

@@

@@

The fitted parameters in comparison with those from men's data are
summarized in Table~\ref{tab:womensfit}.  One notable aspect is that while
the venue effect standard deviation is nearly identical, the women's heat effect
standard deviation is smaller and $\nu$ is much larger.
This suggests that men’s races exhibit greater 
variability in reaction times, possibly due to faster athletes influencing others 
to react more quickly in certain instances. The consistency in the venue effect 
standard deviation across men’s and women’s data indicates that the venue effect 
is not only statistically significant but also consistent in magnitude across 
genders. These findings highlight the potential impact of competition dynamics 
on heat variability and the robustness of venue-level effects.

@@

We can repeat the simulation methods described in the paper to evaluate
the probability of an extreme reaction time.  Table~\ref{tab:Sim_prob_women}
compares the men's and women's results of observing reaction times less than
0.08, 0.09, and 0.1 seconds.  We find that across all three comparisons that
women have a lower probability of having a fast time. The results are slightly
different from those from the men's data, which echoes existing studies
reporting gender differences in RTs \citep[e.g.,][]{lipps2011implications,
babicc2009reaction, panoutsakopoulos2020gender}.

@@

To evaluate a fair reaction time (RT) barrier for women, we compared suggested 
barriers for men and women based on tail probabilities, as shown in 
Table~\ref{tab:Sim_time_women}. The results indicate that men are more likely 
to be disqualified under the current uniform 0.1-second threshold due to their 
generally faster reaction times. This suggests that the same RT standard may not 
have equivalent implications for men and women. Potential adjustments could 
involve raising the barrier for women to align with men’s disqualification rates 
or lowering the barrier for men to match women’s rates. However, further research 
is needed to validate these findings and explore their broader implications. These 
results contribute to ongoing discussions about RT thresholds and emphasize the 
importance of statistical evidence in guiding decisions about competition fairness.

% \section{Results from Data Excluding Positive Disqualified RTs}

% An earlier iteration of the paper fit a model that did not include RTs from
% athletes who were disqualified or did not finish but still registered a RT.  
% However, it was ultimately decided to include these times to better estimate the 
% left tail of the distribution and more accurately predict the probability of a 
% low RT, as described in the main paper.  
% We did not include negative RTs, however, as these represent a mistake 
% of the runner for starting before
% the gun is fired and are thus meaningless in our objective to
% determine a fair RT barrier.  Not all of those disqualified were
% disqualified because of breaking the 0.1 reaction time barrier; there are many
% reasons why an athlete may be disqualified, with the most notable being failed
% drug tests and lane violations.  Nonetheless, in this section,
% we exclude all disqualified RTs to see their effect on the probability of an
% extreme RT.  We fit an identical model to the generalized Gamma model
% presented in Section 3.2 to examine differences results.

% \begin{table}
%   \centering
%   \caption{Probabilities of observing RTs less than threshold 0.08,
%   0.09, and 0.10 seconds based on the
%     fitted GG GAMLSS model with both venue- and heat-level
% random effects.}
%   \begin{tabular}{c c c c}
%    \toprule
%    Data Set & Threshold 0.08 & Threshold 0.09 & Threshold 0.10  \\
%    \midrule
%    Without DQs & $4.93\cdot10^{-5}$ & $3.53\cdot10^{-4}$ &  $1.97\cdot10^{-3}$  \\
%    With DQs & $6.84\cdot10^{-5}$ & $4.95\cdot10^{-4}$ & $2.76\cdot10^{-3}$ \\
%    \bottomrule
%   \end{tabular}
%   \label{tab:DQSim_probability}
% \end{table}

% \jy{Men's data or pooled data?}
% \eds{with or without 2022?}
% \of{I dont know if we have time for this or if this fits the scope of what we
% are working on.}

% Table~\ref{tab:DQSim_probability} shows the effect of removing disqualificated
% times from the analysis.  The probability of observing extreme RTs is lower when
% we remove the 17 observations.
